---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 3 - Due date 02/04/25"
author: "Rachael Stephan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp25.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
#set up chunk options
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2024 **Monthly** Energy Review. Once again you will work only with the following columns: Total Renewable Energy Production and Hydroelectric Power Consumption. Create a data frame structure with these two time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r upload}
#Load/install required package here
library(forecast); library(tseries); library(Kendall); library(cowplot)
library(lubridate); library(tidyverse); library(openxlsx); library(knitr)

#set theme
mytheme <- theme_bw(base_size = 10)+
  theme(axis.title = element_text(size = 10, hjust = 0.5),
        plot.title.position = "panel",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.25),
        plot.caption = element_text(hjust = 0),
        legend.box = "vertical",
        legend.location = "plot",
        axis.gridlines = element_line(color = "grey", linewidth = 0.25),
        axis.ticks = element_line(color = "black", linewidth = 0.5))
theme_set(mytheme)

#upload dataset
renewable_e_prod_consump <- 
  read.xlsx("./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
                                      sheet = "Monthly Data", 
                                      startRow = 13, 
                                      colNames = FALSE)

#get column names
col_units <- 
  read.xlsx("./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
            rows = 11:12,
            sheet="Monthly Data",
            colNames=FALSE)

#set col names
colnames(renewable_e_prod_consump) <- col_units[1,]

#fix dates
renewable_e_prod_consump$Month <- as_date(renewable_e_prod_consump$Month, origin = "1900-01-01")
renewable_e_prod_consump$Month <- paste(month(renewable_e_prod_consump$Month,
                                              label = TRUE, 
                                              abbr = TRUE), 
                                        year(renewable_e_prod_consump$Month))

#select for columns of interest
energy_matrix <- renewable_e_prod_consump %>%
  select(`Month`, 
         `Total Renewable Energy Production`, 
         `Hydroelectric Power Consumption`)

#get first few rows of each column to check structure and values
kable(head(energy_matrix),
      caption = "First few rows of the selected timeseries for analysis")
str(energy_matrix)

#create time series object
energy_ts <- ts(energy_matrix[,2:3],
                start=c(1973,1),
                frequency=12)
```

##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r initial_plots}
#create plots for renewable ts
#time series
energy_ts_plot_renewable <- autoplot(energy_ts[,1], col ="darkgreen")+
  labs(y = paste("Production",col_units[2,5], sep = " "),
       title = "Renewable Energy Production",
       x = "Year")

#ACF
energy_acf_renewable <- Acf(energy_ts[,1],
                                  lag.max=40, 
                                  type="correlation", 
                                  plot=FALSE)

energy_acf_plot_renewable <- autoplot(energy_acf_renewable)+
  labs(title = "Renewable ACF")

#PACF
energy_pacf_renewable <- Pacf(energy_ts[,1],
                                  lag.max=40, 
                                  plot=FALSE)

energy_pacf_plot_renewable <- autoplot(energy_pacf_renewable)+
  labs(title = "Renewable PACF")

#plot the renewable grid
plot_grid(energy_ts_plot_renewable, 
          energy_acf_plot_renewable, 
          energy_pacf_plot_renewable,
          align = "h",
          nrow = 1)

#plot renewable ts
#time series
energy_ts_plot_hydro <- autoplot(energy_ts[,2], col ="blue")+
  labs(y = paste("Consumption",col_units[2,6], sep = " "),
       title = "Hydroelectric Power Consumption")

#ACF
energy_acf_hydro <- Acf(energy_ts[,2],
                        lag.max=40, 
                        type="correlation", 
                        plot=FALSE)

energy_acf_plot_hydro <- autoplot(energy_acf_hydro)+
  labs(title = "Hydroelectric ACF")

#PACF
energy_pacf_hydro <- Pacf(energy_ts[,2],
                          lag.max=40, 
                          plot=FALSE)

energy_pacf_plot_hydro <- autoplot(energy_pacf_hydro)+
  labs(title = "Hydroelectric PACF")

#plot the hydro plot grid
plot_grid(energy_ts_plot_hydro, 
          energy_acf_plot_hydro, 
          energy_pacf_plot_hydro,
          align = "h",
          nrow = 1)
```

### Q2
*From the plot in Q1, do the series Total Renewable Energy Production and Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?*

The renewable energy consumption autocorrelation plots begins with a high magnitude and positive autocorrelation that decreases at a consistent rate with each lag. Both of these timeseries still maintain a high magnitude of correlation at 40 lags. The autocorrelation plot for the hydroelectric energy production shows different behaviour. This plot oscillates between higher and lower magnitude correlations, alternating between positive and negative correlations. This could indicate the presence of seasonality. As the lags progress, the maximum magnitudes of the positive autocorrelation decreases but the maximum magnitude of the negative autocorrelations increases.

All of the partial autocorrelation plots show similar trends. There is a general decay in amplitude of all of the PACF plots. Only the highest amplitudes surpass the blue dotted line that indicates significance. The amount of times that exceed the blue line decreases with increasing lags. There is a smaller oscillating motion between positive and negative PACF that may coincide with seasonality in the data. The renewable energy and biomass energy PACF plots have similar magnitudes, and the hydroelectric energy PACF magnitudes were slightly less in comparison.

### Q3
Use the *lm()* function to fit a linear trend to the two time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r lm_renewable}
#create variable for number of temporal observations
time <- 1:nrow(energy_ts)

#Fit the linear trend for renewable energy
lm_renewable <- lm(energy_ts[,1] ~ time)
summary(lm_renewable)

#save renewable lm coefficients
lm_renewable_beta0 <- as.numeric(lm_renewable$coefficients[1])
lm_renewable_beta1 <- as.numeric(lm_renewable$coefficients[2])
```

***answer renewable here***

```{r lm_hydro}
#Fit the linear trend for hydro energy
lm_hydro <- lm(energy_ts[,2] ~ time)
summary(lm_hydro)

#save renewable lm coefficients
lm_hydro_beta0 <- as.numeric(lm_hydro$coefficients[1])
lm_hydro_beta1 <- as.numeric(lm_hydro$coefficients[2])
```

***answer hydro here*** 
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r detrend_renewable}
#create renewable linear trend equation
lm_trend_renewable <- lm_renewable_beta0 + lm_renewable_beta1 * time

#use linear trend to detrend renewable ts
lm_detrend_renewable <- energy_ts[,1] - lm_trend_renewable


#create timeseries trend
ts_trend_renewable <- ts(lm_trend_renewable,
                         start=c(1973,1),
                         frequency=12)

ts_detrend_renewable <- ts(lm_detrend_renewable,
                           start = c(1973,1),
                           frequency = 12)

#plot ts and trends
autoplot(energy_ts[,1],color="darkblue")+
  autolayer(ts_detrend_renewable, series ="Detrended",color="green")+
  autolayer(ts_trend_renewable, series ="Linear Component",color="red")+
  labs(title = "Trended and Detrended Renewable Energy Production in the USA",
       y = paste("Renewable Energy Production",col_units[2,6], sep = " "),
       x= "Year")
```



```{r detrend_hydro}
#create hydro linear trend equation
lm_trend_hydro <- lm_hydro_beta0 + lm_hydro_beta1 * time

#use linear trend to detrend hydro ts
lm_detrend_hydro <- energy_ts[,2] - lm_trend_hydro


#create timeseries trend
ts_trend_hydro <- ts(lm_trend_hydro,
                         start=c(1973,1),
                         frequency=12)

ts_detrend_hydro <- ts(lm_detrend_hydro,
                           start = c(1973,1),
                           frequency = 12)

#plot ts and trends
autoplot(energy_ts[,2],color="darkblue")+
  autolayer(ts_detrend_hydro, series ="Detrended",color="green")+
  autolayer(ts_trend_hydro, series ="Linear Component",color="red")+
  labs(title = "Trended and Detrended Hydrodroelectric Energy Production in the USA",
       y = paste("Hydroelectric Energy Production",col_units[2,6], sep = " "),
       x= "Year")
```


### Q5
*Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side, but not mandatory. Did the plots change? How?*

```{r detrend_acf_pacf}
#determine ACF, PACF for renewable energy
#ACF
acf_detrend_renewable <- Acf(ts_detrend_renewable,
                  lag.max=40, 
                  type="correlation", 
                  plot=FALSE)

acf_detrend_plot_renewable <- autoplot(acf_detrend_renewable)+
  labs(title = "Detrended Renewable ACF")

#PACF
pacf_detrend_renewable <- Pacf(ts_detrend_renewable,
                  lag.max=40, 
                  plot=FALSE)

pacf_detrend_plot_renewable <- autoplot(pacf_detrend_renewable)+
  labs(title = "Detrended Renewable ACF")

#determine ACF, PACF for hydro energy
#ACF
acf_detrend_hydro <- Acf(ts_detrend_hydro,
                  lag.max=40, 
                  type="correlation", 
                  plot=FALSE)

acf_detrend_plot_hydro <- autoplot(acf_detrend_hydro)+
  labs(title = "Detrended Hydroelectric ACF")

#PACF
pacf_detrend_hydro <- Pacf(ts_detrend_hydro,
                  lag.max=40, 
                  plot=FALSE)

pacf_detrend_plot_hydro <- autoplot(pacf_detrend_hydro)+
  labs(title = "Detrended Hydroelectric ACF")

#plot ACF comparisons in a plot
plot_grid(energy_acf_plot_renewable, 
          acf_detrend_plot_renewable,
          energy_acf_plot_hydro,
          acf_detrend_plot_hydro,
          align = "h",
          nrow = 2)
```

***talk about acf here***

```{r pacf_plots}
#Plot PACF comparisons in a grid
plot_grid(energy_acf_plot_renewable, 
          pacf_detrend_plot_renewable,
          energy_acf_plot_hydro,
          acf_detrend_plot_hydro,
          align = "h",
          nrow = 2)
```

***talk about pacf here***

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r seasonal_lm}
#Use seasonal means model on the trended renewable dataset
ts_dummies_renewable <- seasonaldummy(energy_ts[,1])

lm_seasonal_renewable <- lm(energy_ts[,1] ~ ts_dummies_renewable)
summary(lm_seasonal_renewable)

lm_seasonal_beta_int_renewable <-lm_seasonal_renewable$coefficients[1]
lm_seasonal_beta_coeff_renewable <-lm_seasonal_renewable$coefficients[2:12]

#Use seasonal means model on the trended hydro dataset
ts_dummies_hydro <- seasonaldummy(energy_ts[,2])

lm_seasonal_hydro <- lm(energy_ts[,2] ~ ts_dummies_hydro)
summary(lm_seasonal_hydro)

lm_seasonal_beta_int_hydro <-lm_seasonal_hydro$coefficients[1]
lm_seasonal_beta_coeff_hydro <-lm_seasonal_hydro$coefficients[2:12]
```


### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r deseason_renewable}
seasonal_comp_renewable <- array(0, nrow(energy_ts))


for(i in time){
  seasonal_comp_renewable[i] <- lm_seasonal_beta_int_renewable + 
    lm_seasonal_beta_coeff_renewable %*% ts_dummies_renewable[i,]
}

lm_deseasoned_renewable <- energy_ts[,1] - seasonal_comp_renewable

ts_deseason_renewable <- ts(lm_deseasoned_renewable,
                              start=c(1973,1),
                              frequency = 12)

#plot ts and deseasoned data
autoplot(energy_ts[,1],color="darkblue")+
  autolayer(ts_deseason_renewable, series ="Deseasoned",color="green")+
  labs(title = "Seasoned and Deseasoned Renewable Energy Production in the USA",
       y = paste("Renewable Energy Production",col_units[2,6], sep = " "),
       x= "Year")
```

```{r deseason_hydro}
seasonal_comp_hydro <- array(0, nrow(energy_ts))


for(i in time){
  seasonal_comp_hydro[i] <- lm_seasonal_beta_int_hydro + 
    lm_seasonal_beta_coeff_hydro %*% ts_dummies_hydro[i,]
}

lm_deseasoned_hydro <- energy_ts[,2] - seasonal_comp_hydro

ts_deseason_hydro <- ts(lm_deseasoned_hydro,
                              start=c(1973,1),
                              frequency = 12)

#plot ts and deseasoned data
autoplot(energy_ts[,2],color="darkblue")+
  autolayer(ts_deseason_hydro, series ="Deseasoned",color="green")+
  labs(title = "Seasoned and Deseasoned Hydro Energy Production in the USA",
       y = paste("Hydro Energy Production",col_units[2,6], sep = " "),
       x= "Year")
```

### Q9
*Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side, but not mandatory. Did the plots change? How?*
```{r deseason_acf_pacf}
#determine ACF, PACF for renewable energy
#ACF
acf_deseason_renewable <- Acf(ts_deseason_renewable,
                  lag.max=40, 
                  type="correlation", 
                  plot=FALSE)

acf_deseason_plot_renewable <- autoplot(acf_deseason_renewable)+
  labs(title = "Deseasoned Renewable ACF")

#PACF
pacf_deseason_renewable <- Pacf(ts_deseason_renewable,
                  lag.max=40, 
                  plot=FALSE)

pacf_deseason_plot_renewable <- autoplot(pacf_deseason_renewable)+
  labs(title = "Deseasoned Renewable ACF")

#determine ACF, PACF for hydro energy
#ACF
acf_deseason_hydro <- Acf(ts_deseason_hydro,
                  lag.max=40, 
                  type="correlation", 
                  plot=FALSE)

acf_deseason_plot_hydro <- autoplot(acf_deseason_hydro)+
  labs(title = "Deseasoned Hydroelectric ACF")

#PACF
pacf_deseason_hydro <- Pacf(ts_deseason_hydro,
                  lag.max=40, 
                  plot=FALSE)

pacf_deseason_plot_hydro <- autoplot(pacf_deseason_hydro)+
  labs(title = "Deseasoned Hydroelectric ACF")

#plot ACF comparisons in a plot
plot_grid(energy_acf_plot_renewable, 
          acf_deseason_plot_renewable,
          energy_acf_plot_hydro,
          acf_deseason_plot_hydro,
          align = "h",
          nrow = 2)
```

***talk about acf here***

```{r deseason_pacf_plots}
#Plot PACF comparisons in a grid
plot_grid(energy_pacf_plot_renewable, 
          pacf_deseason_plot_renewable,
          energy_pacf_plot_hydro,
          pacf_deseason_plot_hydro,
          align = "h",
          nrow = 2)
```

***talk about pacf here***

